COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-09 paper/bare_conf.tex
Rephrase vector-register tiling + minor fixes

% Padding and Data Alignment
\subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. In this example, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Data alignment and padding are for an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
\label{code:helmholtz-licm}
\end{algorithm}

@@

\begin{algorithm}[t]
\label{algo:PyOP2Compiler}
  \textbf{The COFFEE Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
\If{isa.backend == gpu}{
  \If{it\_space}{ast.extract\_iteration\_space(wrapper)}
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
// Optimize ast based on plan \\
ast.licm() \\
ast.padding() \\
ast.data\_align() \\
\If{plan.permute}{
  ast.permute\_assembly\_loops()
}
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  uaj = MIN(plan.uaj\_factor, $\lfloor$it\_space.j.size/isa.vf$\rfloor$) \\
  ast.op\_vect(uaj)
}
\Return{wrapper + ast.from\_ast\_to\_c()}
\caption{Pseudocode of the COFFEE pipeline.}
\end{algorithm}

@@

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}. The second transformation step is applied if the backend is a manycore architecture, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}. Then, an ordered sequence of optimization steps are executed. Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, \emph{split}, and \emph{op-vect} may be introduced. Their implementation is based on analysis and transformation of the kernel's AST. When \emph{op-vect} is selected, the compiler outputs proper AVX or LRBNi intrinsics code. Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

@@

Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism, so tuning critical parameters, like the unroll factor, is of great importance. However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters. This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride. We therefore leave the backend compiler in charge of selecting the unroll factor. This choice also simplifies the COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added. Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).

@@

\begin{algorithm}[t]
\label{algo:applyCostModel}
  \textbf{Cost Model}\\
  \KwIn{n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs}
  \KwOut{uaj\_factor, split\_factor}

@@

// Compute unroll-and-jam factor for \emph{op-vect} \\
n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)\\
uaj\_factor = $\lceil$n\_reg\_avail / n\_inner\_arrays$\rceil$ \\
\eIf{n\_outer\_arrays $>$ n\_inner\_arrays}{
  permute = \KwSty{True}
}{
  permute = \KwSty{False}
}
\Return{$<$permute, split\_factor, uaj\_factor$>$}
\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
\end{algorithm}

@@

Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi. The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

@@

Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code. We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment. Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized. This motivates the usually significant gain obtained. Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization. Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned. Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g. in Burgers 3D p3). This is due to the large number of aligned temporaries introduced by \emph{licm}, which probably induces cache associativity conflicts.

\subsection{Impact of Vector-register Tiling}
\label{sec:perf-eval-opvect}
Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels, respectively. For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with the COFFEE's cost model. In general, there is no substantial difference between the two. This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple. The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful. This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short. On the other hand, as the iteration space becomes bigger, vector-register tiling minimizes pipeline stalls, leading to maximum speed-ups of 1.4$\times$ in Diffusion and of 1.24$\times$ in Helmholtz on the Sandy Bridge architecture. This was proved using the Intel Architecture Code Analyzer tool~\cite{IACA}. In Helmholtz, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a consequence of increased register re-use, which relieves the pressure on both data (cache) ports available in the processor.

Application of \emph{op-vect} to the Burgers problem induces meaningful slow downs, due to the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying machine. Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
\hline
Problem & Max slow down & Max speed up \\\hline\hline
Helmholtz & / & 11.72$\times$ \\
Diffusion & / & 14.68$\times$ \\
Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
\end{tabulary}
\end{center}
\label{table:comparison-to-FFC-opt}
\caption{Summary of performance comparison between FEniCS and COFFEE on the Sandy Bridge machine. The slow down and speed up columns compare the best COFFEE implementation and the FEniCS implementation with optimizations enabled.}
\end{table}

\subsection{Comparison with hand-made BLAS-based implementations}
For the Helmholtz 3D problem, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge. This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only \emph{licm}, data alignment and padding were used, with the slow down being even worse for smaller problem sizes. These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance. It is possible that employing certain BLAS implementations or auto-tuning strategies for particular problem sizes is helpful, although the same problems afflicting \emph{split} (exposed in Section~\ref{sec:expr-split}) might limit the performance gain. However, as anticipated, extraction of matrix-matrix multiplies from analysis of the kernel's AST and a strategy for their execution will be comprehensively addressed in further work.

>>>>

% Padding and Data Alignment
\subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying architecture possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. In this example, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Data alignment and padding are for an AVX architecture. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
\label{code:helmholtz-licm}
\end{algorithm}

@@

\begin{algorithm}[t]
  \textbf{The COFFEE Compiler}\\
  \KwIn{ast, wrapper, isa}
  \KwOut{code}
// Analyze ast and build optimization plan \\
it\_space = analyze(ast) \\
\If{\KwSty{not} it\_space}{
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
  \Return{wrapper + ast.from\_ast\_to\_c()}
}
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
// Optimize ast based on plan \\
ast.licm() \\
ast.padding() \\
ast.data\_align() \\
\If{plan.permute}{
  ast.permute\_assembly\_loops()
}
\If{plan.sz\_split}{
  ast.split(plan.sz\_split)
}
\If{plan.uaj\_factor}{
  uaj = MIN(plan.uaj\_factor, $\lfloor$it\_space.j.size/isa.vf$\rfloor$) \\
  ast.op\_vect(uaj)
}
\Return{wrapper + ast.from\_ast\_to\_c()}
\caption{Pseudocode of the COFFEE pipeline.}
\label{algo:PyOP2Compiler}
\end{algorithm}

@@

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}. Then, an ordered sequence of optimization steps are executed. Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, \emph{split}, and \emph{op-vect} may be introduced. Their implementation is based on analysis and transformation of the kernel's AST. When \emph{op-vect} is selected, the compiler outputs proper AVX or LRBNi intrinsics code. Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

@@

Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism, so tuning critical parameters, like the unroll factor, is of great importance. However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters. This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride. We therefore leave the backend compiler in charge of selecting the unroll factor. This choice also simplifies the COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added. Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).

@@

\begin{algorithm}[t]
  \textbf{Cost Model}\\
  \KwIn{n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs}
  \KwOut{uaj\_factor, split\_factor}

@@

// Compute unroll-and-jam factor for \emph{op-vect} \\
n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)\\
uaj\_factor = $\lceil$n\_reg\_avail / n\_inner\_arrays$\rceil$ \\
\eIf{n\_outer\_arrays $>$ n\_inner\_arrays}{
  permute = \KwSty{True}
}{
  permute = \KwSty{False}
}
\Return{$<$permute, split\_factor, uaj\_factor$>$}
\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
\label{algo:applyCostModel}
\end{algorithm}

@@

Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi. The \texttt{icc 13.1} compiler was used. On the Sandy Bridge, the compilation flags used were \texttt{-O2} (other optimization levels perfomed, in general, slightly worse) and \texttt{-xAVX} for auto-vectorization. On the Xeon Phi, the optimization level \texttt{-O3} was used. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

@@

Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Xeon Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code. We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment. Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized. This motivates the usually significant gain obtained. Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization. Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned. Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g. in Burgers 3D p3). This is due to the large number of aligned temporaries introduced by \emph{licm}, which probably induces cache associativity conflicts.

\subsection{Impact of Vector-register Tiling}
\label{sec:perf-eval-opvect}
Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels, respectively. For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with the COFFEE's cost model. In general, there is no substantial difference between the two. This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple.

The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful. This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short. On the other hand, as the iteration space becomes bigger, vector-register tiling leads to percentage improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to X$\times$ on the Xeon Phi (X, X, X). Using the Intel Architecture Code Analyzer tool~\cite{IACA}, we proved that this is a consequence of increased register re-use. In Helmholtz, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.

On the Sandy Bridge, we have also measured the performance of individual kernels in terms of floating-point operations per second. The theoretical peak, with the Intel Turbo Boost technology activated, is 30.4 GFlop/s. In the case of Diffusion using a prismatic mesh and $p=4$, we achived a maximum of 21.9 GFlop/s with \emph{op-vect} enabled, whereas 16.4 GFlop/s was obtained when only \emph{licm-ap} is used. This result is inline with the expectations: analysis of assembly code showed that, in the \texttt{jk} loop nest, which in this problem represents the bulk of the computation, 73$\%$ of instructions are actually floating-point operations.

Application of \emph{op-vect} to the Burgers problem induces meaningful slow downs, due to the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying architecture. Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
\hline
Problem & Max slow down & Max speed up \\\hline\hline
Helmholtz & / & 11.72$\times$ \\
Diffusion & / & 14.68$\times$ \\
Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
\end{tabulary}
\end{center}
\label{table:comparison-to-FFC-opt}
\caption{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}
\end{table}

\subsection{Comparison with hand-made BLAS-based implementations}
For the Helmholtz 3D problem, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge. This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only \emph{licm}, data alignment and padding were used, with the slow down being even worse for smaller problem sizes. These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance. It is possible that employing certain BLAS implementations or auto-tuning strategies, particularly for certain problem sizes, is helpful, although the same problems afflicting \emph{split}, exposed in Section~\ref{sec:expr-split}, might limit the performance gain. However, as anticipated, the problem of identifying and extracting matrix-matrix multiplies from analysis of the kernel's AST, as well as a strategy for their execution, will be comprehensively addressed in further work.
